The effect of prenatal environment on brain metabolic function and perception in a lizard

### 2025-05-29 ms.qmd
updates

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. However, early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects. Here, we investigated the combined influence of prenatal corticosterone (CORT) — a key stress hormone in reptiles — and incubation temperature on mitochondrial function and prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta — two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta but not the olfactory bulbs. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms linking early-life stress, metabolic function, and cognitive performance.

@@

Cognition encompasses the ways in which animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to changing environments [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination of both [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. However, the impact of early-life environments have primarily been studied in the context of learning and memory [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020], neglecting how other equally important cognitive domains might be affected [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. In particular, perception is an often overlooked, but fundamental skill [@shettleworth]. Perceiving different stimuli is essential for locating food, avoiding predators, or interacting with conspecifics [@burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Variations in the ability to detect or recognize relevant stimuli can have significant consequences for an individual’s fitness [@pyke1984optimal]. For example, visual acuity in the bluegill sunfish (_Lepomis macrochirus_) allow animals to detect prey at large distances, which can increase their foraging efficiency and energy intake [@hairston1982fish]. Therefore, understanding how prenatal conditions affect the ability to detect prey can be important to understanding the broader consequences of early-life environments on fitness.

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental inputs during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Long-lasting effects on energy production and oxidative stress in the brain may be a key mechanism by which cognitive function is impacted by developmental environments  [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is expected to depend on efficient mitochondrial respiration. Additionally, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks correlate negatively with the number of round-shaped mitochondria — associated with higher oxidative stress — per presynaptic terminal. These findings emphasize the role of mitochondrial action in shaping cognitive abilities. However, how the early environment influences mitochondrial physiology and its associated cognitive consequences remains largely unknown outside of mammals [see @chaudhari2022early for a review]. In particular, whether and how these interactions affect perceptual abilities, such as prey detection, is still largely unexplored.

**Maternal conditions can play a crucial role in shaping mitochondrial function in the offspring. Because mitochondria are maternally inherited, a mother’s physiological state can directly impact mitochondria through trans-generational effects [@zhu_prenatal_2004; @song2009prenatal; @picard_mitochondria_2014]. One key factor in this process is the action of glucocorticoids (GCs), hormones that regulate energy metabolism and stress responses [@sapolsky_how_2000]. Under stressful situations, animals instigate adaptive physiological and behavioural adjustments that involve increases in GCs. These GCs can be transmitted to the offspring and influence their mitochondrial functioning [@picard_mitochondria_2014]. For instance, prenatal stress or elevated yolk GCs increase oxidative stress in Sprague-Dawley rats (_Rattus norvegicus_) and domestic chickens (_Gallus domesticus_) [@zhu_prenatal_2004; @song2009prenatal; @haussmann_embryonic_2012].**

**In ectotherms, thermal stress experienced by mothers can increase maternal circulating glucocorticoids. Activation of the physiological stress response and higher GC concentrations can subsequently impact offspring [see @Crino_2023]. However, temperature is also known to directly affect offspring development beyond maternal effects, particularly during early life [@noble_developmental_2018; @crino2024eggs]. The prenatal thermal environment can play a critical role in shaping mitochondrial function, altering energy metabolism or oxidative stress [@stier2022experimental; @crino2024eggs]. For example, Red-eared slider turtle (_Trachemys scripta elegans_) hatchlings incubated at 26.5 ºC exhibited greater oxidative damage compared to those incubated at temperatures 0.7 to 1.2 ºC warmer [@treidel2016temperature]. Therefore, the combined effects of prenatal GCs and temperature can have profound consequences for metabolic function, with significant repercussions for brain development and cognition. Yet, we have little evidence of how the interactive effects of prenatal GCs and temperature can impact cognitive abilities by developmental programming of metabolic pathways.**

Here, we investigated the combined effects of prenatal temperature and corticosterone (CORT) - the main GC in reptiles - on mitochondrial function and prey detection in a lizard (_Lampropholis delicata_). We manipulated egg CORT levels and incubation in a fully factorial design. We then assessed hatchling prey detection ability and quantified metabolic function in brain regions related to processing chemical and visual cues. We hypothesized that CORT would impair prey detection through changes in metabolic function by decreasing respiration efficiency [@gong_chronic_2011;  but see @crino2024eggs] and increasing oxidative stress [reviewed in @costantini2011meta]. **In contrast, we predicted that low incubation temperatures would only enhance behavioural performance if the benefits of mitochondrial efficiency outweigh the costs of higher oxidative stress [@treidel2016temperature; @crino2024eggs]. We also expected complex interactions between CORT and temperature, as they have opposing effects on mitochondrial efficiency. Finally, we predicted region-dependent effects of early-life conditions on mitochondrial function [@coomber_independent_1997] that would lead to stimulus-specific differences in prey detection.**

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them under one of two temperature regimes (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either: a) 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or b) an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study where CORT treatment increased mean yolk CORT levels **~3.7x higher than control eggs** [@crino2024eggs]. Eggs were then incubated at either cold  (23 ± 3 ºC) or hot (28 ± 3 ºC) incubation temperatures in a 2x2 factorial design (Fig. 1 A). These temperatures are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig-cap: "Scheme of our experimental design. In panel A, we show the different stages of our experiment and the main manipulations. In panel B, we show the experimental device used to present the stimuli in the behavioural tests. In panel C, we show the experimental setup for the prey discrimination tests. In panel D, we show the relevant times from our behavioural tests."
#
knitr::include_graphics("./Others/Methods.png")
```

@@

_Oxidative Damage_: Assays of oxidative damage from cryopreserved samples were performed **within 6 months** of the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples, removed the cryopreservation solution (see above), and resuspended samples in 1 mL warm 1X Tris-EDTA. Afterward, the Tris-EDTA was removed, and samples were resuspended in 200 µL of warm 1X PBS containing 10 µg/mL Hoechst 33342 Nuclear Viability Dye (for cell viability) and 100 µM BODIPY 665/676 Lipid Peroxidation Sensor (to measure lipid peroxidation). Following staining, we permeabilized the cell membranes incubating the samples in 200 µL warm 1X PBS containing 20 µM digitonin. Following permeabilization, we stained the samples with 20 µL of 70 µM 8-OHdG Polyclonal Antibody to measure oxidative damage on DNA. We left the homogenate overnight (~12 hours) at 4 ºC. The following day we counterstained the cells with 20 µL of 100 µg/mL H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyze the samples in the flow cytometer.

@@

We obtained direct, indirect, and total effects from posterior distributions of parameters estimated in the multivariate models. Direct effects represent the posterior estimates of a predictor’s effect on the response variable, **while indirect effects were calculated by multiplying these effects on a mediator by the mediator’s effect on the response** [@kline2005principles]. Total effects are the sum of direct and indirect effects.

@@

## Discussion
In this study, we examined the impact of prenatal temperature and CORT elevations on lizards' ability to detect different stimuli from prey together with the impacts of early conditions on mitochondrial metabolic function and oxidative stress in two brain regions associated to perception of visual and chemical stimuli. Our findings suggest that mitochondrial metabolic function remained largely resilient to early environmental conditions, yet oxidative damage and behavioral responses exhibited region-specific and context-dependent variations. Below, we discuss these findings in relation to mitochondrial metabolic function, oxidative stress, and behavioral performance.

#### Mitochondrial function and ROS production is robust to early life conditions
We found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in either of the brain regions studied. This suggests that, contrary to our predictions, mitochondrial function is robust to prenatal temperature and CORT. Previous studies have shown that high incubation temperatures - but not CORT - decreased mitochondrial efficiency in the liver of _L. delicata_ [@crino2024eggs]. Mitochondrial potential is expected to enhance ATP production per oxygen consumed, thereby increasing metabolic efficiency [@rice2002brain; @martinez2016tca]. Given the brain's high energy demands [@mcnay_decreases_2000; @mann_coupling_2021], it is plausible that mitochondria are particularly well-protected, which may help explain their resilience towards environmental inputs. Future studies should explore how the environment influences mitochondrial function accross tissues, as well as the mechanisms underlying the brain’s resilience to early-life conditions.

#### The interactive effects of prenatal CORT and temperature on oxidative damage are region-dependent
We found that DNA damage and lipid peroxidation were affected in region-specific ways. Early conditions did not significantly influence oxidative stress in the olfactory bulbs, but had a pronounced effect on DNA damage and lipid peroxidation in the optic tecta.

CORT exposure reduced DNA damage in optic tecta from cold-incubated lizards, but had no effect at higher incubation temperatures. This pattern suggests that CORT might play a protective role in specific thermal conditions. Since CORT helps mobilize energy resources to facilitate physiological adjustments to stress [@sapolsky_how_2000], it is possible that under colder incubation conditions — where metabolic rates are expected to be lower — these resources can be directed towards antioxidant activity or DNA repairing mechanisms. In contrast, at higher temperatures, the increased metabolic demands may limit the availability of resources for such protective mechanisms.

In comparison, prenatal CORT elevations increased lipid peroxidation in cold incubated lizards but reduced it in those incubated at higher temperatures. Although the two hormonal treatments did not significantly differ within each incubation temperature, the direction of the CORT effect was the opposite between temperatures. One possible reason for this pattern could be temperature-dependent changes in membrane lipid composition or fluidity, which may affect how lipids respond to oxidative stress [@zeis2019temperature]. For instance, lipid peroxidation levels increased in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids [@zeis2019temperature]. Future studies should explore how temperature-dependent effects on membrane composition influence oxidative damage susceptibility.

#### Detection of chemical, but not visual, stimulus improved with prenatal CORT exposure
Detection latency decreased with CORT exposure when animals were presented with a chemical stimulus, but not with a visual stimulus. This effect was not related to alterations in mitochondrial physiology in the olfactory bulbs. Instead, we hypothesize that it may be driven by differences in motivation. Although lizards were fasted for two days before testing and we specifically tested for motivation, we may have overlooked factors related to the the ecological relevance of the cues, the animals' internal state, or the interaction between both factors. Prenatal CORT is expected to increase hunger through resource mobilization [@spencer2008post; @cossin-sevrin_effect_2022; but see @crino_corticosterone_2014-learn]. This could have lead to increase motivation in CORT-treated lizards, which could have switched to a more active foraging mode, affecting the way they respond to chemical stimuli. For example, species of European lizards (family Lacertidae) that spend more time actively looking for food tend to rely more on chemoreception [@verwaijen2007relationships]. Further studies need to be conducted to understand how CORT can change foraging strategies in lizards and how this can affect their perception abilities.

#### Mitochondrial function and oxidative stress do not affect detection abilities
Despite the relevance of mitochondrial function and oxidative damage on cognitive abilities in other taxa [@zhu_prenatal_2004; @cao2014ampk; @hara_presynaptic_2014] we found no significant relationships between mitochondrial physiology and prey detection in _L. delicata_. This suggest that the brain regions studied are not directly related to the behaviors tested, or that other mechanisms are more relevant for preceiving prey stimuli. Alternatively, mitochondrial physiology may influence detection abilities only beyond a certain threshold, which may not have been reached in our study.

In this regard, mitochondria are predicted to impact cognitive performance by providing energy for neural activity or by regulating oxidative stress [@siegel1994basic; @mcnay_decreases_2000; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018; @mann_coupling_2021; @alexandrov_neuronal_2022]. However, the lack of effect of mitochondrial density and metabolic capacity on detection latency abilities in _L. delicata_ suggests that energy production is not a limiting factor for prey perception, or that _L. delicata_ can sustain this cognitive function despite fluctuations in energy availability. Similarly, the levels of damage in the brain regions studied may not have been high enough to impair prey perception in this species or compensatory mechanisms may have counteracted any negative effects of oxidative stress.

Despite not seeing any effect of oxidative damage, since DNA damage or lipid peroxidation can accumulate over time, their effects on cognition can pronounce over time [@terman2006oxidative; @hara_presynaptic_2014]. In fact, we saw a significant effect of age on DNA damage in olfactory bulbs, and on lipid peroxidation in both regions. Future research should investigate the long-term consequences of mitochondrial activity on brain health and the potential mechanisms that sustain cognitive resilience in _L. delicata_.

#### Conclusions
Prenatal incubation and CORT exposure had region-specific effects in the brain of _L. delicata_. While mitochondrial function remained resilient to early environmental conditions, oxidative damage and behavioral responses exhibited context-dependent variations. Our findings suggest that the brain regions studied are not directly related to the behaviors tested, or that other factors, like motivation, are more relevant for preceiving prey stimuli. Future studies should explore how the environment influences mitochondrial function accross tissues, as well as the mechanisms underlying the brain’s resilience to early-life conditions.

>>>>

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. However, early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects. Here, we investigated the combined influence of prenatal corticosterone (CORT) — a key stress hormone in reptiles — and incubation temperature on mitochondrial function and prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta — two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta but not the olfactory bulbs. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms linking early-life stress, cell metabolism, and cognitive performance.

@@

Cognition encompasses the ways in which animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to changing environments [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination of both [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. However, the impact of early-life environments has primarily been studied in the context of learning and memory [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020], neglecting how other equally important cognitive domains might be affected [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. In particular, perception is an often overlooked, but fundamental skill [@shettleworth]. Perceiving different stimuli is essential for locating food, avoiding predators, or interacting with conspecifics [@burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Variations in the ability to detect or recognize relevant stimuli can have significant consequences for an individual’s fitness [@pyke1984optimal]. For example, visual acuity in the bluegill sunfish (_Lepomis macrochirus_) allow animals to detect prey at large distances, which can increase their foraging efficiency and energy intake [@hairston1982fish]. Therefore, understanding how prenatal conditions affect the ability to detect prey can be important to understanding the broader consequences of early-life environments on fitness.

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental inputs during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Long-lasting effects on energy production and oxidative stress in the brain may be a key mechanism by which cognitive function is impacted by developmental environments [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is expected to depend on efficient mitochondrial respiration. Additionally, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks correlate negatively with the number of round-shaped mitochondria — associated with higher oxidative stress — per presynaptic terminal. In conclusion, findings from different studies suggest that mitocondrial activity can be essential in shaping cognitive abilities. Thus, understanding how early-life conditions affect mitochondrial physiology may provide insights into the mechanisms underlying cognitive performance.

Mitochondria are highly responsive to environmental conditions during early development, with potential short-term and long-lasting consequences [reviewed in @gyllenhammer2020developmental]. For example, environmental cues such as external conditions or hormone exposure can cause immediate disruptions to metabolic function, triggering cascading effects on oxidative stress that may persist leter in life [reviewed in @gyllenhammer2020developmental]. Among the range of environmental factors that can shape mitochondrial physiology, some exert particularly strong and consistent effects. Stress-related hormones and temperature are particularly relevant for mitochondrial function due to their strong effects on metabolic processes [@sapolsky_how_2000; @stier2022experimental; @crino2024eggs]. Under stressful situations, animals instigate adaptive physiological and behavioural adjustments that involve increases in glucocorticoids (GCs), which are also important metabolic regulators [@sapolsky_how_2000]. Similarly, temperature plays a prominent role in regulating metabolic activity in ectotherms [@stier2022experimental; @crino2024eggs]. In fact, GC elevations or changes in the thermal environment during early development have profound and sustained effects on mitochondrial respiration efficiency or oxidative stress [@zhu_prenatal_2004; @song2009prenatal; @costantini2011meta; @haussmann_embryonic_2012; @treidel2016temperature; @stier2022experimental; @crino2024eggs]. However, how the early environment influences mitochondrial physiology and its associated cognitive consequences remains largely unknown outside of mammals [see @chaudhari2022early for a review]. In particular, whether and how these interactions affect perceptual abilities, such as prey detection, is still largely unexplored.

Here, we investigated the combined effects of prenatal temperature and corticosterone (CORT) - the main GC in reptiles - on mitochondrial physiology and prey detection in a lizard (_Lampropholis delicata_). We manipulated egg CORT levels and incubation in a fully factorial design. We then assessed hatchling prey detection ability and quantified metabolic function in brain regions related to processing chemical and visual cues. We hypothesized that prenatal CORT exposure and the incubation thermal environment would influence metabolic function and oxidative stress in the brain [@zhu_prenatal_2004; @song2009prenatal; @costantini2011meta; @haussmann_embryonic_2012; @treidel2016temperature; @stier2022experimental; @crino2024eggs], with significant repercussions on cognition [@zhu_prenatal_2004; @hara_presynaptic_2014]. We predicted that metabolic deficiencies or increases in oxidative stress and damage would result in lower cognitive abilities (i.e. longer time to detect prey) [@picard_energetic_2018; @alexandrov_neuronal_2022]. Furthermore, we predicted prenatal conditions to have region-dependent effects on mitochondrial function [@coomber_independent_1997] that would lead to stimulus-specific differences in prey detection.

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them under one of two temperature regimes (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either: a) 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or b) an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study where CORT treatment increased mean yolk CORT levels by approximately 2 standard deviations above the mean natural concentration [@crino2024eggs]. Eggs were then incubated at either cold  (23 ± 3 ºC) or hot (28 ± 3 ºC) incubation temperatures in a 2x2 factorial design (Fig. 1 A). These temperatures are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig-cap: "Scheme of our experimental design. In panel A, we show the different stages of our experiment and the main manipulations. In panel B, we show the experimental device used to present the stimuli in the behavioural tests. Here, F indicates the front of the device, and Bk the back. In panel C, we show the experimental setup for the prey discrimination tests. In panel D, we show the relevant times from our behavioural tests."
#
knitr::include_graphics("./Others/Methods.png")
```

@@

_Oxidative Damage_: Assays of oxidative damage from cryopreserved samples were performed **5 months** after the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples, removed the cryopreservation solution (see above), and resuspended samples in 1 mL warm 1X Tris-EDTA. Afterward, the Tris-EDTA was removed, and samples were resuspended in 200 µL of warm 1X PBS containing 10 µg/mL Hoechst 33342 Nuclear Viability Dye (for cell viability) and 100 µM BODIPY 665/676 Lipid Peroxidation Sensor (to measure lipid peroxidation). Following staining, we permeabilized the cell membranes incubating the samples in 200 µL warm 1X PBS containing 20 µM digitonin. Following permeabilization, we stained the samples with 20 µL of 70 µM 8-OHdG Polyclonal Antibody to measure oxidative damage on DNA. We left the homogenate overnight (~12 hours) at 4 ºC. The following day we counterstained the cells with 20 µL of 100 µg/mL H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyze the samples in the flow cytometer.

@@

We obtained direct, indirect, and total effects from posterior distributions of parameters estimated in the multivariate models. Direct effects represent the posterior estimates of a predictor’s effect on the response variable, while indirect effects were computed as the product of the direct effects along the path [@kline2005principles]. Total effects are the sum of direct and indirect effects.

@@

## Discussion
We examined the impact of prenatal temperature and CORT on visual and chemical perception and determined whether metabolic function was the underlying reason for any perceptual differences we observed as has been shown in mammals [@hara_presynaptic_2014; @chaudhari2022early]. Our findings suggest that metabolic processes remained largely resilient to early environmental conditions. However, early-life conditions have region- and context-dependent effects on oxidative damage and prey detection. We observed interactive effects of prenatal CORT and incubation temperature on oxidative damage in the Optic Tecta, but no effects on perception of visual stimuli. In contrast, CORT decrease the time to perceived chemical stimuli despite the resilience of the Olfactory bulbs to prenatal conditions. Below, we discuss these findings in relation to mitochondrial metabolic function, oxidative stress, and behavioural performance.

#### Metabolic processes in multiple brain regions are robust to early life conditions
Contrary to our predictions, we found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in either of the brain regions studied. **EXAMPLES OF OTHER EXPERIMENTS WITH TEMP AND CORT IN THE BRAIN.** Mitochondrial potential is expected to enhance ATP production per oxygen consumed, thereby increasing metabolic efficiency [@rice2002brain; @martinez2016tca]. Given the brain's high energy demands [@mcnay_decreases_2000; @mann_coupling_2021], it is plausible that mitochondria are particularly well-protected, which may help explain their resilience towards environmental inputs.

Alternatively, the effects of early environmental conditions on metabolic function and ROS production may be immediate rather than long-lasting. Mitochondria are known to be highly responsive to environmental changes and energetic demands [@gyllenhammer2020developmental]. It is possible that the metabolic effect of the early-life conditions dissipated once all animals were exposed to the same environment after hatching. Nonetheless, alterations in metabolic function during earlier stages would still be expected to affect oxidative damage, as these processes are cumulative and can lead to long-term consequences [@rice2002brain; @terman2006oxidative].

#### Oxidative stress brought about by experiencing early life CORT are region- and temperature- dependent
We found that DNA damage and lipid peroxidation were affected in region-specific ways. Early conditions did not significantly influence oxidative stress in the olfactory bulbs, but had a pronounced effect on DNA damage and lipid peroxidation in the optic tecta.

CORT exposure reduced DNA damage in optic tecta from cold-incubated lizards but had no effect at higher incubation temperatures. This pattern suggests that CORT might play a protective role under specific thermal conditions. Since CORT helps mobilize energy resources to facilitate physiological adjustments to stress [@sapolsky_how_2000], it is possible that under colder incubation conditions — where metabolic rates are expected to be lower — energy mobilized by CORT may be allocated towards antioxidant activity or DNA repairing mechanisms [**REFs**]. In contrast, at higher temperatures, the increased metabolic demands may limit the availability of resources for such protective mechanisms [**REFs**]. **A further possibility is that, under lower metabolic activity induced by cold temperatures more portion of DNA is protected within nucleosomes/histones, decreasing the effects of CORT on oxidative damage** [**REFs**].

In comparison, increased prenatal CORT resulted in higher lipid peroxidation in cold incubated lizards but reduced it in those incubated at warm temperatures. Although the two hormonal treatments did not significantly differ within each incubation temperature, the direction of the CORT effect was the opposite between temperatures. The opposing temperature-dependent effects of CORT on DNA damage and lipid peroxidation indicates that the pathways involved in protection or repair are target-specific. **EXAMPLE OF ENZYMES WORKING ON LIPID PEROXIDATION AND DNA DAMAGE AT DIFFERENT TEMPERATURES**. An alternative mechanism might involve temperature-dependent changes in membrane lipid composition or fluidity, which may affect how lipids respond to oxidative stress [@zeis2019temperature]. For instance, lipid peroxidation levels increased in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids [@zeis2019temperature]. In that case, temperature-induced structural differences may influence the vulnerability of different molecules to oxidative stress.

#### Prenatal CORT exposure enhances chemical but not visual prey detection   
Detection latency decreased with CORT exposure when animals were presented with chemical, but not visual, stimuli. This effect was not related to alterations in mitochondrial physiology in the olfactory bulbs. Instead, prenatal CORT may influence other mechanisms involved in chemoreception. For instance, in rats (_Rattus norvegicus_), GCs like dexamethasone can enchance the ability to detect chemical stimuli by upregulating gene expression in the olfactory mucosa [@meunier2020olfactory]. Such molecular changes could alter chemosensory sensitivity in lizards, independent of mitochondrial function.

Alternatively, the observed differences in chemical perception may be driven by motivation. Prenatal CORT is expected to increase hunger through resource mobilization [@spencer2008post; @cossin-sevrin_effect_2022; but see @crino_corticosterone_2014-learn]. This could have led to increased motivation in CORT-treated lizards, stimulating a switch to a more active foraging mode, affecting the way they respond to chemical stimuli. For example, species of European lizards (family Lacertidae) that spend more time actively looking for food tend to rely more on chemoreception [@verwaijen2007relationships]. Although we fasted lizards and controlled for motivation in our analyses, it is possible that we did not fully capture factors related to the ecological relevance of the cues, the animals' internal state, or the interaction between the two. Further studies need to be conducted to understand how CORT can change foraging strategies in lizards and how this can affect their perception abilities.

#### Mitochondrial function and oxidative stress do not affect detection abilities
Mitochondria are predicted to impact cognitive performance by providing energy for neural activity or by regulating oxidative stress [@siegel1994basic; @mcnay_decreases_2000; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018; @mann_coupling_2021; @alexandrov_neuronal_2022]. However, despite the relevance of metabolic function and oxidative damage on cognitive abilities in other taxa [@zhu_prenatal_2004; @cao2014ampk; @hara_presynaptic_2014] they did not explain variation in prey detection in _L. delicata_. Metabolic function and oxidative damage did not have substantial effects on prey detection in _L. delicata_.

Previous studies have found that **EXAMPLE ABOUT MIT DENSITY OR SIMILAR AND PREY DETECTION IF I CAN FIND ANY BECAUSE DAN DOES NOT REALIZE THAT NOONE DOES THIS**. In contrast, our results suggests that metabolic function is not a limiting factor for prey detection in _L. delicata_, or that _L. delicata_ can sustain this cognitive function despite fluctuations in energy availability.

Similarly, oxidative stress did not influence _L. delicata_ ability to detect visual or chemical stimuli from prey. In this regard, oxidative damage is expected to decrease cognitive abilities through neuron death or senescense [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Possibly the levels of oxidative damage observed in our study were not sufficient to impair prey detection either because of compensatory mechanisms - such as antioxidant activity - that counteracted the negative effects of oxidative stress, or because the studied brain regions had a high neural density that provided functional resilience [@rice2002brain; @amiel_smart_2011]. Alternatively, the cognitive effects of oxidative stress may not yet be detectable at this stage. Oxidative damage accumulates over time, and its effects on cognition can pronounce with age [@terman2006oxidative; @hara_presynaptic_2014]. In fact, we saw a significant effect of age on DNA damage in olfactory bulbs, and on lipid peroxidation in both regions. Cross-sectional studies across life stages ay help reveal how oxidative damage shapes cognitive function over time.

#### Conclusions
Prenatal incubation and CORT exposure had region-specific effects in the brain of _L. delicata_. While mitochondrial function remained resilient to early environmental conditions, oxidative damage and behavioural responses exhibited context-dependent variations. Our findings suggest that the brain regions studied are not directly related to the behaviours tested, or that other factors, like motivation, are more relevant for detecting prey stimuli. Future studies should explore how the environment influences mitochondrial function accross tissues, as well as the mechanisms underlying the brain’s resilience to early-life conditions.
